---
title: "In-Class_Ex4"
date: '9 December 2023'
date-modified: 'last-modified'
format: html
execute: 
  eval: true # run the code live
  echo: true # all code will appear
  warning: false # hide all warnings
editor: visual
---

## Getting Started

```{r}
pacman::p_load(tidyverse, sf, httr, tmap, performance)
```

## Geocoding using SLA API

```{r}
url <- 'https://www.onemap.gov.sg/api/common/elastic/search'

csv <- read_csv('data/aspatial/Generalinformationofschools.csv')
postcodes <- csv$`postal_code`

found <- data.frame()
not_found <- data.frame()

for(postcode in postcodes) {
  query <- list('searchVal' = postcode,
                'returnGeom' = 'Y',
                'getAddrDetails' = 'Y',
                'pageNum' = '1')
  res <- GET(url,query = query)
  if((content(res)$found)!= 0 ){
    found <- rbind(found, data.frame(content(res))[4:13])
  }
  else{
    not_found = data.frame(postcode)
  }
}
```

Next, the code chunk below will be used to combine both *found* and *not_found* data frames into a single tibble df called merged. At the same time, we will write *merged* and *not_found* tibble dfs into csv format for future use.

```{r}
#| eval: false
merged <- merge(csv, found, by.x = 'postal_code', by.y = 'results.POSTAL', all = TRUE)

write.csv(merged, file = 'data/aspatial/schools.csv')

write.csv(not_found, file = 'data/aspatial/not_found.csv')
```

## Converting an aspatial data frame into a simple feature tibble data frame

### Importing and tidying *schools.csv* data frame

```{r}
schools <- read_csv('data/aspatial/schools.csv')

schools <- schools %>%
  rename(latitude = results.LATITUDE,
         longitude = results.LONGITUDE) %>%
  select(postal_code, school_name, latitude, longitude)
```

### Converting an aspatial data frame into sf tibble data frame

```{r}
schools <- st_as_sf(schools, 
                    coords = c('longitude','latitude'),
                    crs = 4326)%>%
  st_transform(crs = 3414)
```

### Plotting a point simple feature layer

To ensure that *schools* sf df has been projected and converted correctly, you can plot the school points data for visual inspection.

```{r}
tmap_mode('view')

tm_shape(schools)+
  tm_dots()+
  tm_view(set.zoom.limits = c(11,14))
```

## Preparing

### Point-in-Polygon Count

Importing *Master Planning Sub-zone 2019*

```{r}
mpsz <- st_read(dsn = 'data/geospatial',
                layer = 'MPSZ-2019') %>%
  st_transform(crs = 3414)
```

Count number of school that falls within a planning subzone.

```{r}
mpsz$`SCHOOL_COUNT` <- lengths(st_intersects(mpsz, schools))
```

### Importing Business shapefile

```{r}
business <- st_read(dsn = 'data/geospatial',
                    layer = 'Business')
```

```{r}
tmap_options(check.and.fix = TRUE)
tmap_mode('plot')

tm_shape(mpsz)+
  tm_polygons()+
  tm_shape(business)+
  tm_dots()
```

Create a new column to count the number of Businesses in each mpsz.

```{r}
mpsz$`business_count` <- lengths(st_intersects(mpsz, business))
```

```{r}
summary(mpsz$business_count)
```

## The Data

```{r}
flow_data <- read_rds('data/rds/flow_data_tidy.rds')
glimpse(flow_data)
```

Let's check for 0 values.

```{r}
summary(flow_data)
```

```{r}
flow_data$FlowNoIntra <- ifelse(
  flow_data$ORIGIN_SZ == flow_data$DESTIN_SZ,
  0, flow_data$MORNING_PEAK)
flow_data$offset <- ifelse(
  flow_data$ORIGIN_SZ == flow_data$DESTIN_SZ,
  0.000001, 1)

inter_zonal_flow <- flow_data %>% 
  filter(FlowNoIntra >0)

inter_zonal_flow <- inter_zonal_flow %>% 
  rename(TRIPS = MORNING_PEAK,
         DIST = dist)
```

### Origin (Production) Constrained SIM

```{r}
orcSIM_Poisson <- glm(formula = TRIPS ~
                        ORIGIN_SZ +
                        log(SCHOOL_COUNT)+
                        log(RETAIL_COUNT)+
                        log(DIST) - 1, #the -1 is to remove the intercept which is not necessary in a constrained model
                      family = poisson(link = 'log'),
                      data = inter_zonal_flow,
                      na.action = na.exclude)

summary(orcSIM_Poisson)
```

#### Goodness-of-Fit

```{r}
CalcRSquared <- function(observed,estimated){
  r <- cor(observed, estimated)
  R2 <- r^2
  R2
}
```

We can examine how the constraints hold for destinations this time.

```{r}
CalcRSquared(orcSIM_Poisson$data$TRIPS, orcSIM_Poisson$fitted.values)
```

```{r}
performance_rmse(orcSIM_Poisson,
                 normalized = FALSE)
```

### Doubly Constrained

```{r}
dbcSIM_Poisson <- glm(formula = TRIPS ~
                        ORIGIN_SZ +
                        DESTIN_SZ +
                        log(DIST), # No - 1 after log(DIST)
                      family = poisson(link = 'log'),
                      data = inter_zonal_flow,
                      na.action = na.exclude)
```

```{r}
CalcRSquared(dbcSIM_Poisson$data$TRIPS, dbcSIM_Poisson$fitted.values)
```

```{r}
performance_rmse(dbcSIM_Poisson,
                 normalized = FALSE)
```
